Seabird assemblages are linked to the major western boundary current off eastern Australia

### 2023-10-24 ms_preprint/ms_preprint.tex
ms adjustments after LB RS and MS reviews pre-submission

\author{
    Nicholas W. Daudt*
   \\
    Department of Marine Science \\
    University of Otago \\
  Dunedin, Aotearoa New Zealand \\
  \texttt{\href{mailto:nicholaswdaudt@gmail.com}{\nolinkurl{nicholaswdaudt@gmail.com}}} \\
   \And
    Eric J. Woehler
   \\
    Australasian Seabird Group \\
    BirdLife Australia \\
  Hobart, Australia \\
  \texttt{\href{mailto:eric.woehler@gmail.com}{\nolinkurl{eric.woehler@gmail.com}}} \\
   \And
    Matthew R. Schofield
   \\
    Department of Mathematics and Statistics \\
    University of Otago \\
  Dunedin, Aotearoa New Zealand \\
  \texttt{\href{mailto:matthew.schofield@otago.ac.nz}{\nolinkurl{matthew.schofield@otago.ac.nz}}} \\
   \And
    Robert O. Smith
   \\
    Department of Marine Science \\
    University of Otago \\
  Dunedin, Aotearoa New Zealand \\
  \texttt{\href{mailto:robert.smith@otago.ac.nz}{\nolinkurl{robert.smith@otago.ac.nz}}} \\
   \And
    Leandro Bugoni
   \\
    Institute of Biological Sciences \\
    Universidade Federal do Rio Grande--FURG \\
  Rio Grande, RS, Brazil \\
  \texttt{\href{mailto:lbugoni@yahoo.com.br}{\nolinkurl{lbugoni@yahoo.com.br}}} \\
   \And
    William J. Rayment
   \\
    Department of Marine Science \\
    University of Otago \\
  Dunedin, Aotearoa New Zealand \\
  \texttt{\href{mailto:will.rayment@otago.ac.nz}{\nolinkurl{will.rayment@otago.ac.nz}}} \\
  }

@@

\begin{abstract}
\textbf{Aim:} To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
\textbf{Location:} East coast of Australia.
\textbf{Methods:} We used 15 seabird at-sea ship-based survey data sets collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.
\textbf{Results:} Based on 142,646 seabirds recorded, from 80 species, all models suggested two macro-scale assemblages (`northern' and `southern'), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at \textasciitilde34°S, near the latitude at which the East Australian Current separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
\textbf{Main conclusions:} RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the rapid climatic changes that the East Australian Current is experiencing and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
\end{abstract}

@@

William J. Rayment -- \url{https://orcid.org/0000-0001-9558-5109}

@@

We thank all observers who assisted in data collection, the crews of RV \emph{Investigator} and RV \emph{Falkor}, and the Australasian Seabird Group (BirdLife Australia) for logistical support. The authors would like to thank Dr.~Skipton N.C. Woolley, \texttt{\{ecomix\}} developer, who kindly helped NWD build preliminary models and gave advice in several stages of model construction. NWD is supported by a Doctoral scholarship from the University of Otago.

@@

\textbf{Data availability statement:} The gridded data used as the basis for all analyses are available in the archived repository along with the code to reproduce the results. The raw data are currently being used for other projects and thus can not be publicly available. If you are interested in the raw data, please get in touch with Eric J. Woehler.

@@

\textbf{Results:} Based on 142,646 seabirds recorded, from 80 species, all models suggested two macro-scale assemblages (`northern' and `southern'), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at \textasciitilde34°S, near the latitude at which the East Australian Current separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.

\textbf{Main conclusions:} RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the rapid climatic changes that the East Australian Current is experiencing and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.

@@

Seabird assemblages have been widely described and related to environmental characteristics (Hyrenbach et al., 2007; Ribic et al., 1997; Woehler et al., 2003). Considering the hierarchical nature of oceanographic processes and their resulting features (Haury et al., 1978), it is not surprising that the distribution of highly mobile seabirds are linked differently across temporal and spatial scales (Haury et al., 1978; Hunt \& Schneider, 1987). In the marine environment, large scale (\textgreater100 km length) oceanographic features (e.g.~boundary currents and water masses) act as terrestrial biomes (based on vegetation), and influence seabird assemblage structure--i.e.~the occurrence of species and their relative numerical composition (Commins et al., 2014; Hyrenbach et al., 2007; Ribic et al., 1997). In contrast, local abundance is related to meso- (10--100 km) and sub-mesoscale (1--10 km) processes that aggregate prey and therefore predators, such as fronts and eddies (Scales, Miller, Hawkes, et al., 2014; Weimerskirch, 2007). Such processes, although relatively ephemeral, are primarily driven by physical and atmospheric forces, and are thus somewhat predictable in space and time, providing consistent feeding areas for marine predators when they occur (Davoren, 2013; Hazen, Suryan, et al., 2013). Sea surface temperatures are a key driver of seabird distributions, at individual- and assemblage-levels, operating over a range of spatial and temporal scales from macro (e.g.~water masses) to fine (e.g.~fronts) (Dunlop et al., 1988; Quillfeldt et al., 2015; Ribic et al., 1997). Despite extensive surveys (e.g. Ballance et al., 1997) and tracking studies (e.g. Carneiro et al., 2020), many oceanic regions still lack information on the distributions and abundances of seabird species, and how they assemble in space and time.

@@

In this study, we aim to identify assemblages of seabird species within the EAC and adjacent areas, and relate these assemblages to environmental conditions, by using at-sea survey data and a state-of-the-art modelling technique. We use the Region of Common Profiles {[}RCP; Foster et al. (2013){]}, a mixture model regarded as one of the best available tools for describing bioregions (N. A. Hill et al., 2020; S. N. C. Woolley et al., 2020). RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics (N. A. Hill et al., 2020; Warton et al., 2015). Eastern Australia supports a high diversity of seabirds and offers wide environmental gradients over a near 40° latitudinal range, from the tropics to the cool temperate. We hypothesised to find different seabird assemblages relating to different environmental conditions within the EAC. In addition, presence-absence and abundance data were used to test if the different numerical resolutions would result in the identification of similar assemblages. This study provides a repeatable framework on which future studies can assess how future changes in the EAC may influence the distribution of seabird assemblages.

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\textasciitilde18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC (Ganachaud et al., 2014; Kessler \& Cravatte, 2013). The South Caledonian Jet completes the EAC system at \textasciitilde22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (\textbf{Figure \ref{fig:fig1-study-area}a}; Kessler \& Cravatte (2013); Ganachaud et al. (2014)). The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies (Oke, Roughan, et al., 2019; Oke \& Middleton, 2000; Ridgway \& Dunn, 2003). The EAC flows polewards and is primarily associated with the shelf break, but at \textasciitilde32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea (Mata et al., 2006; Oke, Roughan, et al., 2019), the `EAC eastern extension' (Oke, Pilo, et al. (2019); often termed the `Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the `EAC southern extension'; \textbf{Figure \ref{fig:fig1-study-area}a}), carrying eddy-like features as a transient current (Oke, Roughan, et al., 2019).

@@

Seabird data were gathered aboard \emph{RV Investigator} and \emph{RV Falkor} on 15 voyages, between 2016/17--2020/21, spanning over \textasciitilde37° of latitude (\textbf{Figure \ref{fig:fig1-study-area}b}; \textbf{Table S1}). One or more observers continuously recorded seabirds within a 300 m forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours while the ship was underway. The species, the numbers of birds, and their behaviours were logged in real time into a ship-based web server. Environmental parameters for atmospheric and oceanographic conditions were logged continuously by automated-ship based sensors, with seabird observations supplementing relevant environmental conditions such as visibility or associations with marine mammals. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species). Birds following the vessels were excluded from analyses. Seabird observation protocols followed Raymond et al. (2010), in use around Australia since 1980.

To investigate and predict species' distributions and relationships with environmental variables, a considered choice of broad-scale climatic and oceanographic covariates is important (Barbet-Massin \& Jetz, 2014). We used three physiographic variables and seven oceanographic variables, from which a further three climatic layers were calculated (\textbf{Table \ref{tab:table1}}). These variables are the most frequently used in seabird-environment relationship studies (Tremblay et al., 2009), capturing processes at a range of spatial and temporal scales from macro-scale water-mass properties to prey availability in meso- to submesoscale oceanographic features (Haury et al., 1978; Scales, Miller, Hawkes, et al., 2014; Weimerskirch, 2007).

@@

Data wrangling, modelling and visualization were done in R 4.2.0 (R Core Team, 2022); the full list of packages can be found in the Supplementary Material. Code and the modelling dataset (see \ref{sec:modellingdatasets}) are available in \textbf{(Zenodo repo)}.

@@

The spatial and temporal persistence of oceanographic features such as eddy kinetic energy (`eke') and thermal fronts (sea surface temperature gradient: `sst\_grad') may influence the movements and distributions of seabirds by offering a higher likelihood of finding food (Scales, Miller, Embling, et al., 2014; Weimerskirch, 2007). Thus, `eke' and `sst\_grad' environmental layers were calculated before extracting environmental data for each seabird record. For both `eke' and `sst\_grad', we used their original cell spatial resolution (\textbf{Table \ref{tab:table1}}) and calculations were done layer-by-layer to match seabird data on the finest spatial scale possible. Eddy kinetic energy was calculated as \(0.5 (u'² + v'²)\), where \(u' = u - \bar{u}\) and \(v' = v - \bar{v}\) are the surface zonal (latitudinal) and meridian (longitudinal) geostrophic currents, respectively, using a rolling mean of 90 days as \(\bar{u}\) and \(\bar{v}\) (Sun et al., 2022). The spatial `sst\_grad' for each cell was calculated by averaging the maximum value for each neighbouring cell (average neighbourhood maximum), from north-to-south (NS) and west-to-east (WE); then, the magnitude of the gradient was defined as \(NS² + WE²\) (Burrows et al., 2011). In addition, we calculated climatic seasonal average layers for `eke' and `sst\_grad' based on the 11-year data for 2011--2021 as a proxy for persistent oceanographic features, and the `eke' standard deviation over the same period (named with the prefix `clim\_' in \textbf{Table \ref{tab:table1}}). Seabed slope was calculated using `raster::terrain()' (Hijmans, 2022a), and distance from the coast was calculated through `sf::st\_distance()' (Pebesma, 2018) using Australia and Tasmania as the mainland (i.e.~ignoring smaller islands). Then, for each seabird record, the most proximate value for each environmental layer (original or calculated) was extracted using built-in functions from `raster' 3.5-21, `terra' 1.6-7, and `sf' 1.0-8 packages (Hijmans, 2022a, 2022b; Pebesma, 2018).

@@

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods (Schreiber \& Burger, 2002). Therefore, seasonal differences in species composition were expected (e.g. K. Reid et al., 2023), and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling datasets for each season. We aggregated seabird data by 1° latitude \(\times\) 1° longitude grid cells (\emph{c.} 110 \(\times\) 110 km) and season (\textbf{Figure \ref{fig:fig1-study-area}b}), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \textgreater1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

We used the mixture model Region of Common Profiles (RCP) to identify and describe seabird species assemblages. RCP is a one-stage bioregionalization approach, as it identifies bioregions by specifying a statistical model that simultaneously relates response (multi-species data) and predictor (environmental data) (N. A. Hill et al., 2020; S. N. C. Woolley et al., 2020). Given a set of environmental variables, the model can predict assemblage probabilities for sites where no biotic (observational) data exist. As a model-based approach, RCPs can be compared by information criteria, present uncertainty around the results, and check for model assumptions through assessing distribution of residuals (N. A. Hill et al., 2020; Warton et al., 2015).

The model is specified in terms of assemblages and site-specific probabilities. There are \(K\) assemblages (i.e.~RCPs), \(K = (1, ..., k)\), that are common across the study area. The outcome variable is a vector \(y_{i} = (y_{i1}, ... y_{iS})\), that gives the count or presence-absence of each of the \(S\) species at site \(i\). Depending on whether we model the counts or presence-absence data, the assemblages are defined in terms of the expected frequency or relative frequency of the species, respectively. An output of the model is the estimated probability that site \(i\) has each of the \(K\) possible assemblages. RCPs are a multivariate adaptation of a mixture-of-experts model; as such, the mixing proportions (assemblage structure) vary according to covariates (predictors) (Foster et al., 2013). However, the final estimated assemblage structure is assumed to be constant for all sites belonging to the \(K\) assemblages (Foster et al., 2013). As such, RCPs can be directly interpreted as assemblages, and their species profiles as the assemblage structure.

@@

Due to seabird migration and dispersion, seasons could act as `sampling artifacts' (Foster et al., 2017), thus biasing results (Ponti \& Sannolo, 2023). In light of the difference among seasons in the numbers of species that met our occurrence threshold (see \ref{sec:modellingdatasets}), we decided to use RCP models by season and data type (presence-absence and abundance). RCP models were used to relate the response variables (i.e.~presence-absence or abundance of each seabird species in each grid cell) to the independent predictor covariates (i.e.~averaged values of environmental variables in each cell). Presence-absence data were modelled based on a Bernoulli model, whereas abundance data were based on a Negative Binomial model. RCPs were fitted using the `ecomix' 1.0.0 package (S. Woolley et al., 2022), with the following workflow: (i) find the best number of groups (\(K\)) that describes the data, (ii) re-fit the model by fixing \(K\) and selecting the optimal combination of covariates, (iii) check residuals, (iv) bootstrap the best model to estimate uncertainty around parameter estimates, and (v) produce results.

@@

A total of 10,261 seabird records was obtained during the 15 voyages, comprising 142,646 individuals from 80 species, belonging to 11 families and five orders. The five most abundant species made up 83\% of total birds counted (43.0\% short-tailed shearwater \emph{Ardenna tenuirostris}, 17.2\% fairy prion \emph{Pachyptila turtur}, 14.3\% wedge-tailed shearwater \emph{Ardenna pacifica}, 5.2\% sooty tern \emph{Onychoprion fuscatus}, 3.2\% shy albatross \emph{Thalassarche cauta}). The five most frequently encountered species were \emph{A. pacifica} (frequency of occurrence = 35.3\%), \emph{T. cauta} (33.2\%), brown booby \emph{Sula leucogaster} (25.4\%), red-footed booby \emph{S. sula} (25.4\%) and black-browed albatross \emph{T. melanophris} (25\%) (\textbf{Figure S1}). An increase in species richness towards the south of the study region was clear in all but the winter season (\textbf{Figure S2}). Noticeably higher counts were made around Tasmania, possibly due to the `colony effect' and/or the presence of Southern Ocean species (\textbf{Figure S2}). However, one cell in the Coral Sea had counts of the same order of magnitude as off Tasmania in summer.

@@

@@

The gridded data used as the basis for all analyses are available in the archived repository along with the code to reproduce the results. The raw data are currently being used for other projects and thus can not be publicly available. If you are interested in the raw data, please get in touch with XXX.

>>>>

\author{
    Nicholas W. Daudt
    \thanks{Corresponding author}
   \\
    Department of Marine Science \\
    University of Otago \\
  Dunedin, Aotearoa New Zealand \\
  \texttt{\href{mailto:nicholaswdaudt@gmail.com}{\nolinkurl{nicholaswdaudt@gmail.com}}} \\
   \And
    Eric J. Woehler
   \\
    Australasian Seabird Group \\
    BirdLife Australia \\
  Hobart, Australia \\
  \texttt{\href{mailto:eric.woehler@gmail.com}{\nolinkurl{eric.woehler@gmail.com}}} \\
   \And
    Matthew R. Schofield
   \\
    Department of Mathematics and Statistics \\
    University of Otago \\
  Dunedin, Aotearoa New Zealand \\
  \texttt{\href{mailto:matthew.schofield@otago.ac.nz}{\nolinkurl{matthew.schofield@otago.ac.nz}}} \\
   \And
    Robert O. Smith
   \\
    Department of Marine Science \\
    University of Otago \\
  Dunedin, Aotearoa New Zealand \\
  \texttt{\href{mailto:robert.smith@otago.ac.nz}{\nolinkurl{robert.smith@otago.ac.nz}}} \\
   \And
    Leandro Bugoni
   \\
    Institute of Biological Sciences \\
    Universidade Federal do Rio Grande---FURG \\
  Rio Grande, RS, Brazil \\
  \texttt{\href{mailto:lbugoni@yahoo.com.br}{\nolinkurl{lbugoni@yahoo.com.br}}} \\
   \And
    William J. Rayment
   \\
    Department of Marine Science \\
    University of Otago \\
  Dunedin, Aotearoa New Zealand \\
  \texttt{\href{mailto:will.rayment@otago.ac.nz}{\nolinkurl{will.rayment@otago.ac.nz}}} \\
  }

@@

\begin{abstract}
\textbf{Aim:} To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
\textbf{Location:} East coast of Australia.
\textbf{Methods:} We used 15 seabird at-sea ship-based survey data sets collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.
\textbf{Results:} Results were based on 142,646 seabirds recorded, from 80 species, including albatrosses, petrels, prions, shearwaters, boobies, and terns, among other taxa. All models suggested two macro-scale assemblages (`northern' and `southern'), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at \textasciitilde34°S, near the latitude at which the East Australian Current (EAC) separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
\textbf{Main conclusions:} RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given ongoing poleward intensification that the EAC is experiencing, which is projected to continue over the next century, and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
\end{abstract}

@@

William J. Rayment \url{https://orcid.org/0000-0001-9558-5109}

@@

We thank all observers who assisted in data collection, the crews of RV \emph{Investigator} and RV \emph{Falkor}, and the Australasian Seabird Group (BirdLife Australia) for logistical support. The authors would like to thank Dr.~Skipton N.C. Woolley, \texttt{\{ecomix\}} developer, who kindly helped NWD build preliminary models and gave advice in several stages of model construction. NWD is supported by a Doctoral scholarship from the University of Otago; LB is a research fellow from CNPq (310145/2022-8).

@@

\textbf{Data availability statement:} The gridded data used as the basis for all analyses are available in the archived repository along with the code to reproduce the results. The seabird raw data are currently being used for other projects and thus can not be publicly available. If you are interested in the seabird raw data, please contact Eric J. Woehler. The environmental data used in this study are freely available from the source (see Table 1 in the main text).

@@

\textbf{Results:} Results were based on 142,646 seabirds recorded, from 80 species, including albatrosses, petrels, prions, shearwaters, boobies, and terns, among other taxa. All models suggested two macro-scale assemblages (`northern' and `southern'), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at \textasciitilde34°S, near the latitude at which the East Australian Current (EAC) separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.

\textbf{Main conclusions:} RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given ongoing poleward intensification that the EAC is experiencing, which is projected to continue over the next century, and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.

@@

Seabird assemblages have been widely described and related to environmental characteristics (Hyrenbach et al., 2007; Ribic et al., 1997; Woehler et al., 2003). Considering the hierarchical nature of oceanographic processes and their resulting features (Haury et al., 1978), it is not surprising that the distribution of highly mobile seabirds are linked differently across temporal and spatial scales (Haury et al., 1978; Hunt \& Schneider, 1987). In the marine environment, large-scale oceanographic features (e.g.~boundary currents and water masses) act similarly to terrestrial biomes (based on vegetation), and influence seabird assemblage structure---i.e.~the occurrence of species and their relative numerical composition (Commins et al., 2014; Hyrenbach et al., 2007; Ribic et al., 1997). In contrast, local abundance is related to meso- (10--100 km) and sub-mesoscale (1--10 km) processes that aggregate prey and therefore predators, such as fronts and eddies (Scales, Miller, Hawkes, et al., 2014; Weimerskirch, 2007). Such processes, although relatively ephemeral, are primarily driven by physical and atmospheric forces, and are thus somewhat predictable in space and time, providing consistent feeding areas for marine predators when they occur (Davoren, 2013; Hazen, Suryan, et al., 2013). Sea surface temperatures are a key driver of seabird distributions, at individual- and assemblage-levels, operating over a range of spatial and temporal scales from macro (e.g.~water masses) to fine (e.g.~fronts) (Dunlop et al., 1988; Quillfeldt et al., 2015; Ribic et al., 1997). Despite extensive surveys (e.g. Ballance et al., 1997) and tracking studies (e.g. Carneiro et al., 2020), many oceanic regions still lack information on the distributions and abundances of seabird species, and how they assemble in space and time.

@@

In this study, we aim to identify assemblages of seabird species within the EAC and adjacent areas, and relate these assemblages to environmental conditions, by using at-sea survey data and a state-of-the-art modelling technique. We use the Region of Common Profiles -- RCP (Foster et al., 2013), a mixture model capable of describing bioregions (N. A. Hill et al., 2020; S. N. C. Woolley et al., 2020). RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics (N. A. Hill et al., 2020; Warton et al., 2015). Eastern Australia supports a high diversity of seabirds and offers wide environmental gradients over a near 40° latitudinal range, from the tropics to the cool temperate. We hypothesised to find different seabird assemblages relating to different environmental conditions within the EAC. In addition, presence-absence and abundance data were used to test if the different numerical resolutions would result in the identification of similar assemblages. This study provides a repeatable framework on which future studies can assess how future changes in the EAC may influence the distribution of seabird assemblages.

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\textasciitilde18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC (Ganachaud et al., 2014; Kessler \& Cravatte, 2013). The South Caledonian Jet completes the EAC system at \textasciitilde22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (\textbf{Figure} \ref{fig:fig1-study-area}\textbf{a}; Kessler \& Cravatte (2013); Ganachaud et al. (2014)). The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies (Oke, Roughan, et al., 2019; Oke \& Middleton, 2000; Ridgway \& Dunn, 2003). The EAC flows polewards and is primarily associated with the shelf break, but at \textasciitilde32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea (Mata et al., 2006; Oke, Roughan, et al., 2019), the `EAC eastern extension', often termed the `Tasman Front' (Oke, Pilo, et al., 2019). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the `EAC southern extension'; \textbf{Figure} \ref{fig:fig1-study-area}\textbf{a}), carrying eddy-like features as a transient current (Oke, Roughan, et al., 2019).

@@

Seabird data were gathered aboard \emph{RV Investigator} and \emph{RV Falkor} on 15 voyages, between 2016/17--2020/21, spanning over \textasciitilde37° of latitude (\textbf{Figure} \ref{fig:fig1-study-area}\textbf{b}; \textbf{Table S1}). One or more observers continuously recorded seabirds within a 300 m forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours while the ship was underway. The species, the numbers of birds, and their behaviours were logged in real time into a ship-based web server. Environmental parameters for atmospheric and oceanographic conditions were logged continuously by automated-ship based sensors, with seabird observations supplementing relevant environmental conditions such as visibility or associations with marine mammals. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species). Birds following the vessels were excluded from analyses. Seabird observation protocols followed Raymond et al. (2010), in use around Australia since 1980.

To investigate and predict species' distributions and relationships with environmental variables, a considered choice of broad-scale climatic and oceanographic covariates is important (Barbet-Massin \& Jetz, 2014). We used three physiographic variables and seven oceanographic variables, from which a further three climatic layers were calculated (\textbf{Table} \ref{tab:table1}). These variables are the most frequently used in seabird-environment relationship studies (Tremblay et al., 2009), capturing processes at a range of spatial and temporal scales from macro-scale water-mass properties to prey availability in meso- to submesoscale oceanographic features (Haury et al., 1978; Scales, Miller, Hawkes, et al., 2014; Weimerskirch, 2007).

@@

Data wrangling, modelling and visualization were done in R 4.2.0 (R Core Team, 2022); the full list of packages can be found in the Supplementary Material. Code and the modelling dataset (see Section \ref{sec:modellingdatasets}) are available in \textbf{(Zenodo)}.

@@

The spatial and temporal persistence of oceanographic features such as eddy kinetic energy (`eke') and thermal fronts (sea surface temperature gradient: `sst\_grad') may influence the movements and distributions of seabirds by offering a higher likelihood of finding food (Scales, Miller, Embling, et al., 2014; Weimerskirch, 2007). Thus, `eke' and `sst\_grad' environmental layers were calculated before extracting environmental data for each seabird record. For both `eke' and `sst\_grad', we used their original cell spatial resolution (\textbf{Table} \ref{tab:table1}) and calculations were done layer-by-layer to match seabird data on the finest spatial scale possible. Eddy kinetic energy was calculated as \(0.5 (u'^{2} + v'^{2})\), where \(u' = u - \bar{u}\) and \(v' = v - \bar{v}\) are the surface zonal (latitudinal) and meridian (longitudinal) geostrophic currents, respectively, using a rolling mean of 90 days as \(\bar{u}\) and \(\bar{v}\) (Sun et al., 2022). The spatial `sst\_grad' for each cell was calculated by averaging the maximum value for each neighbouring cell (average neighbourhood maximum), from north-to-south (NS) and west-to-east (WE); then, the magnitude of the gradient was defined as \(NS^{2} + WE^{2}\) (Burrows et al., 2011). In addition, we calculated climatic seasonal average layers for `eke' and `sst\_grad' based on the 11-year data for 2011--2021 as a proxy for persistent oceanographic features, and the `eke' standard deviation over the same period (named with the prefix `clim\_' in \textbf{Table} \ref{tab:table1}). Seabed slope was calculated using `raster::terrain()' (Hijmans, 2022a), and distance from the coast was calculated through `sf::st\_distance()' (Pebesma, 2018) using Australia and Tasmania as the mainland (i.e.~ignoring smaller islands). Then, for each seabird record, the most proximate value for each environmental layer (original or calculated) was extracted using built-in functions from `raster' 3.5-21, `terra' 1.6-7, and `sf' 1.0-8 packages (Hijmans, 2022a, 2022b; Pebesma, 2018).

@@

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods (Schreiber \& Burger, 2002). Therefore, seasonal differences in species composition were expected (e.g. K. Reid et al., 2023), and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling datasets for each season. We aggregated seabird data by 1° latitude \(\times\) 1° longitude grid cells (\emph{c.} 110 \(\times\) 110 km) and season (\textbf{Figure} \ref{fig:fig1-study-area}\textbf{b}), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \textgreater1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

We used the mixture model Region of Common Profiles (RCP) to identify and describe seabird species assemblages. RCP is a one-stage bioregionalization approach, as it identifies bioregions by specifying a statistical model that simultaneously relates response (multi-species data) and predictor (environmental data) (N. A. Hill et al., 2020; S. N. C. Woolley et al., 2020). Given a set of environmental variables, the model can predict assemblage probabilities for sites where no biotic (observational) data exist. As a model-based approach, RCPs can be compared by information criteria, present uncertainty around the results, and check for model assumptions through assessing residuals (N. A. Hill et al., 2020; Warton et al., 2015).

The model is specified in terms of assemblages and site-specific probabilities. There are \(K\) assemblages (i.e.~RCPs), \(K = (1, ..., k)\), that are common across the study area. The outcome variable is a vector \(y_{i} = (y_{i1}, ... y_{iS})\), that gives the count or presence-absence of each of the \(S\) species at site \(i\). Depending on whether we model the counts or presence-absence data, the assemblages are defined in terms of the expected frequency or relative frequency of the species, respectively. An output of the model is the estimated probability that site \(i\) has each of the \(K\) possible assemblages. RCPs are a multivariate adaptation of a mixture-of-experts model; as such, the mixing proportions (assemblage structure) vary according to covariates (predictors) (Foster et al., 2013). RCPs can be directly interpreted as assemblages; the assemblage structure is given by the species profiles.

@@

Due to seabird migration and dispersion, seasons could act as `sampling artifacts' (Foster et al., 2017), thus biasing results (Ponti \& Sannolo, 2023). In light of the difference among seasons in the numbers of species that met our occurrence threshold (see Section \ref{sec:modellingdatasets}), we decided to use RCP models by season and data type (presence-absence and abundance). RCP models were used to relate the response variables (i.e.~presence-absence or abundance of each seabird species in each grid cell) to the independent predictor covariates (i.e.~averaged values of environmental variables in each cell). Presence-absence data were modelled based on a Bernoulli model, whereas abundance data were based on a negative binomial model. RCPs were fitted using the `ecomix' 1.0.0 package (S. Woolley et al., 2022), with the following workflow: (i) find the best number of groups (\(K\)) that describes the data, (ii) re-fit the model by fixing \(K\) and selecting the optimal combination of covariates, (iii) check residuals, (iv) bootstrap the best model to estimate uncertainty of parameter estimates, and (v) produce results.

@@

A total of 10,261 seabird records was obtained during the 15 voyages, comprising 142,646 individuals from 80 species, belonging to 11 families and five orders. The five most abundant species made up 83\% of total birds counted (43.0\% short-tailed shearwater \emph{Ardenna tenuirostris}, 17.2\% fairy prion \emph{Pachyptila turtur}, 14.3\% wedge-tailed shearwater \emph{Ardenna pacifica}, 5.2\% sooty tern \emph{Onychoprion fuscatus}, 3.2\% shy albatross \emph{Thalassarche cauta}). The five most frequently encountered species were \emph{A. pacifica} (frequency of occurrence = 35.3\%), \emph{T. cauta} (33.2\%), brown booby \emph{Sula leucogaster} (25.4\%), red-footed booby \emph{S. sula} (25.4\%) and black-browed albatross \emph{T. melanophris} (25\%) (\textbf{Figure S1}). An increase in species richness towards the south of the study region was clear in all but the winter season (\textbf{Figure S2}). Noticeably higher counts were made around Tasmania, possibly due to the `colony effect' and/or the presence of Southern Ocean species (\textbf{Figure S2}). However, in summer, one cell in the Coral Sea had counts of the same order of magnitude as off Tasmania.

@@

\leavevmode\vadjust pre{\hypertarget{ref-huang2021}{}}%
Huang, B., Liu, C., Banzon, V., Freeman, E., Graham, G., Hankins, B., Smith, T., \& Zhang, H.-M. (2021). Improvements of the daily optimum interpolation sea surface temperature (DOISST) version 2.1. \emph{Journal of Climate}, \emph{34}(8), 2923--2939. \url{https://doi.org/10.1175/JCLI-D-20-0166.1}

@@

The gridded data used as the basis for all analyses are available in the archived repository along with the code to reproduce the results. The seabird raw data are currently being used for other projects and thus can not be publicly available. If you are interested in the seabird raw data, please contact XXX. The environmental data used in this study are freely available from the source (see Table 1 in the main text).
